Project Thesis/Report

### 2022-01-06 Gandhi_Project_Report/draft2.tex
Done working through all edits Ch1-Ch3

\chapter{Introduction}

@@

The report is organised as follows: 
\newline In \ednote{M: Please label Sections as Chapters as this is apt for a thesis; for the paper that we will write soon, it would be in Sections. Please note the labelling in the tex file, and you may like it to be cross references like this}Chapter~\ref{Chapter_Ds}, we recall the definition of a discrete-time dynamical system, how a discrete-system arises from a flow of an ODE and then proceed to define the inverse limit space and topological conjugacy of autonomous systems. 
\newline In Section 3, we introduce the problem of forecasting dynamical systems, state the Takens delay embedding theorem and discuss various issues faced while forecasting. 
\newline In Section 4, a driven dynamical system is defined and discuss the properties of a specific class of driven dynamical systems that we make use of in this thesis/report/paper. 
\newline Finally in Section 5 we show the implementation of these forecasting methods, and conclusions are provided in Section 6.

\chapter{Discrete-time Dynamical Systems} \label{Chapter_Ds}

@@

Formally, a function $T: U \to U$, where $U$ is some set is  a \emph{discrete-time dynamical system} and its iterates $\{u,Tu,T^2u,\ldots\}$, where $T^n$ denotes the $n$-fold composition of $T$ with itself, describe the evolution of an initial condition $u\in U$ (Note that we drop the brackets and denote $T(u)$ by $Tu$ frequently for simplicity in notation).

Continuous-time dynamical systems modelled using ordinary differential equations can give rise to discrete-time dynamical systems. To see this, consider a differential equation $\dot{x} = f(x)$, $f: \mathbb{R}^n \to \mathbb{R}^n$, $n\in\mathbb{N}$ given to have a unique solution passing through 
each point $x\in\mathbb{R}^{n}$, the flow of the equation \label{defn_flow} is defined to be a mapping $\varphi: \mathbb{R}^n \times \mathbb{R} \to \mathbb{R}$ that satisfies the following properties \ednote{M: Please change this or revert back to how it was defined earlier. We are describing $\varphi$ and in the first two points, it does not concern $\varphi$}
\vspace{-8mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item $x(t)$ is a solution of the differential equation,
  \item $x(0)=x_0$, and
  \item $\varphi(x_0,t) = x(t)$.
\end{enumerate}
By fixing $t=K \in (0,\infty)$, we can define the \emph{time-$K$ map} as  $T(x):= \varphi(x_0,K)$, and it is easily verified that $T\circ T(x_0) = \varphi(x_0,2K)$. In general the $m^{\mbox{th}}$ iterate of $x_0$ under $T$ would be the value of the solution of the ODE evaluated at time $mK$ with the initial condition $x_0$. 
Thus ordinary differential equations give rise to a discrete-time dynamical system by sampling the value of $x$ at time intervals $K$ units apart.

A numerical discretization of a differential equation can also give rise to a discrete-time dynamical system. For instance, Euler's method approximates $\dot{x}(t)$ by $(x(t+h)-x(t))/h$; if $h$ is fixed throughout, the solution of a differential equation $\dot{x}=f(x)$ 
can be approximated at the time instant $t+(m+1)h$ by iterating the equation 
$$x(t+(m+1)h) = x(t+mh) + h f(x(t+mh)).$$ 
Adopting more succinct notation by replacing $x(t+mh)$ with $u_m$, we rewrite the above equation as
$$u_{m+1} = u_m + hf(u_m)$$
or, in even simpler terms, as the discrete-time dynamical system with map $T(u) = u + hf(u)$, where $T$, $u$ and $f$ are understood to be as above.

Of course, discrete-time dynamical systems need not always arise through a differential equation, and once again we may consider the field of ecology where discrete dynamical systems are often directly derived or assumed. 
There is a school of thought that advocates discrete-time dynamical systems to be more natural for modelling real-world observations than differential equations and we refer the interested reader to \cite{saber2010introduction}. \textbf{too philosphical, or actually good?} \ednote{M: It is good! I have moved this paragraph}

\subsection{Invariant Sets}

A core concept in the study of dynamical systems is that of invariance. Given a discrete-time dynamical system $T: U \to U$, a subset $A \subset U$ is said to be an \emph{invariant set} if $T(A) =A$. 
We also define the \emph{orbit of $T$} to be a sequence $\bar{u} = \{u_n\}_{n\in \mathbb{Z}}$ obeying the update equation, $u_{n+1}=Tu_n$, $n \in \mathbb{Z}$.

Two examples of invariant sets include a fixed point where $Tu=u$ for  $u\in U$, and a periodic orbit, i.e., a set of iterates $\{u,Tu, T^2u,\ldots,T^pu\}$, where $T^{p+1}u=u$ for some $p$.  The entire space $U$ could be also be invariant.
Consider for example the space $U=[0,1]$, where $Tu=4u(1-u)$ and then $U$ is invariant. \emph{(as every $u\in{U}$ can be written as $u = 4x(1-x)$ for some $x\in{U}$)}

@@

However,  when the space $U$ is compact, all orbits approach an invariant set. This is since the set of limit points of the orbit can be shown to be invariant \cite{de2013elements}. \ednote{M: I have edited here} So, when $U$ is compact,  the $\omega$-limit set $\omega(u;T)$ of a point $u$ defined to the collection of limit points of the sequence $\{x,Tu,T^2u,\ldots\}$ is nonempty, and $\omega(u;T)$ is invariant. For instance, for the map $Tu=u^2$ defined on $[0,1]$ all orbits lie in the invariant set $\{1\}$ or else would would approach the invariant set $\{0\}$. 
\ednote{B: not relevant other than being an additional example perhaps? M: Have added another simple example}.

We discuss various properties of invariant sets: Invariant sets can be attracting or repelling depending on how orbits in its vicinity behave. Recall the examples $Tu =u/2$ and $Tu=2u$ defined on $\mathbb{R}$ ($\{0\}$ ``attracts" orbits) or the example, $Tu=u^2$ defined on $[0,1]$ ($\{1\}$ ``repels" orbits). We are interested in attractive invariant sets since they capture the long-term dynamics as time increases, and in particular the invariant sets that are called attractors whose definition we recall next. \begin{Definition} \rm Let $T: U \to U$, where $U$  is a metric space with metric $d$. A compact subset $A \subset U$ is said to be an attractor if it satisfies the three conditions: 
  \vspace{-8mm}
  \begin{enumerate}
	\item $A$ is invariant. 
	\item $A$ is asymptotically stable, i.e., for every $\epsilon > 0$ and for all $u$ so that $d(u,A) < \epsilon$, we have $d(T^nu,A) \to 0$ as $n\to \infty$. 
	\item $A$ has Lyapunov stability, i.e., for every $\epsilon > 0$  there exists a $\delta(\epsilon) > 0$ so that $d(u,A) < \delta$ implies $d(T^nu,A) < \epsilon$ for all $n\ge 0$.  
\end{enumerate}
\end{Definition}

@@

\subsection{Chaos}

\ednote{ should one add paragraphs like these? M: this is a thesis, and not a paper; the instructions sent to the external examiner regarding the thesis evaluation says that the student suggest strongly that it is good to write}
\newline In the 1960s, a number of mathematicians and mathematically interested scientists independently discovered chaos in the mathematical sense. The meteorologist Edward Lorenz may have been the first to explain this phenomenon in his 1963 paper \cite{lorenz1963deterministic}. The notions of invariance, attractivity and chaos may also bedescribed for continuous systems, and Lorenz's system comprised of a system of differential equations. The narrative of Lorenz's discovery of chaos, as well as the history of other forerunners in this subject, are fascinating. For those who are interested, we highly recommend James Gleick's book Chaos: The Making of a New Science \cite{gleick2008chaos}. It clearly illustrates these experiences, explains why chaos was such a startling and crucial mathematical and scientific discovery, and describes the underlying mathematical notions for non-specialists.

There are several intricately different definitions of chaos available in the literature, with each of them indicating some aspect of complexity \ednote{B:  what are we saying here? M: Devaney's definition is not the only one, there are about half a dozen definitions as well} In practice, it is not possible to verify which definition or notion does the underlying dynamical system satisfies when only data is observed from it. We do, however, specifically recall the definition of chaos in the sense of Devaney \cite{devaney2018introduction,de2013elements} so as to understand some nuances behind the complexity. Devaney's definition of chaos has three requirements, with the first being the notion  of sensitive dependence on initial conditions.

\begin{Definition} \rm 
A dynamical system $T: U \to U$ is said to have sensitive dependence on initial conditions (SDIC) if there exists a $\delta > 0$ such that for every $u \in U$ and in every neighborhood of $u \in U$ there exists a $v\in{U}$ and an integer $N:=N{(u,v)}\in\mathbb{Z}$ such that $d(T^Nu,T^Nv)>\delta$. 	
\end{Definition}

It is important to acquire a sense of this definition as it can easily be misunderstood. It is common misconception to interpret SDIC as two close points ($u$ and $v$) that eventually become separated by a distance $\delta$ under iteration by $T$. But this is not true. To fully comprehend all the subtleties of the concept, one needs to discuss it more thoroughly by considering each sentence with care and attention, whereafter one may examine how they fit together to convey the concept of sensitive dependence. 
\ednote{ $\leftarrow$ is this talking too simply perhaps? M: No! it is not so}.

@@

\ednote{M: It is probably best to see a few numerical simulations to understand SDIC. (Please use the logistic map and perhaps the trajectories of the double pendulum data). }

@@

\begin{Definition}
	A dynamical system $F: U \to U$ is topologically transitive if for any pair of nonempty open sets $E_1$ and $E_2$ there exists a $n\in\mathbb{N}$ such that $T^n(E_1) \cap E_2 \not= \emptyset$. 
\end{Definition}

Topologically transitivity implies that the iterates of an open set of initial conditions gets mixed up with other open sets. On a compact metric space, it can be shown that topological transitivity also implies the existence of a point whose forward iterates are dense \cite{de2013elements}; or in other words, the orbit going through this point will be dense in the compact metric space. 
In fact, it is topologically more likely that the choice of an arbitrary point will be one  whose iterates are almost dense. 
% I inserted a definition here: it makes more sense to me. 
\begin{Definition} \rm
A subset of a topological space $U$ is said to be a meager set if it cab be written as a countable union of sets of with empty interior. The complement of the a meager set is set to be a residual set.
\end{Definition}

@@

We may now define the notion of Chaos as formulated by Devaney\cite{devaney2018introduction}.
\begin{Definition} \rm
	A dynamical system $T: U \to U$ is said to exhibit chaos in the sense of Devaney if it satisfies the three properties:
	\vspace{-5mm}
  \begin{enumerate}
		\item $T$ has SDIC.
		\item $T$ is topologically transitive.
		\item The set of periodic points of $T$ are dense in $U$. 
	\end{enumerate}
\end{Definition}

\begin{figure}[ht]
  \centering
  \begin{minipage}{.5\textwidth}
    \centering
\includegraphics[scale=0.18]{chaos1.png}
    \captionof{figure}{Starting to diverge}
    \label{fig:chaos1}
  \end{minipage}%
  \begin{minipage}{.5\textwidth}
    \centering
    \includegraphics[scale=0.18]{chaos2.png}
    \captionof{figure}{After more time-steps}
    \label{fig:chaos2}
  \end{minipage}
  \end{figure}

\ednote{M: Please put this picture in perspective and refer after the definition of SDIC. You can also say that you pick two positions that are slightly different but with the same angular velocity. Also in the text, say that you describe the DP in a later chapter.}
\ednote{M: I realize that need figures in the eps format to have the commutativity diagrams compile as well. }

\begin{Example} \rm
  The standard tent map $Tu=1-|2u-1|$ defined on $[0,1]$ is a well known example of a dynamical system satisfying these three properties. 
\end{Example}

  \begin{figure}[ht]
\includegraphics[scale=0.6]{T2.png}
    \centering
    \label{fig:T2tentmap}
  \end{figure}

Let us now consider why this map in the above example exhibits chaos in the sense of Devaneys. The graph of the map $T$ is piecewise linear with two straight lines, one connecting the points $(0,0)$ and $(1/2,1)$ and the other connecting $(1/2,1)$ with $(1,0)$. They form a so-called tent with base centered at $u=1/2$. The graph of the map $T^2$ comprises two symmetric tents with their base centered at $1/4$ and $3/4$.
  % We now reason as why this map exhibits chaos in the sense of Devaney.

  In general, the graph of  $T^k$ contains $2^{k-1}$ tents.
Given any point $u$ and an open neighborhood $W \subset[0,1]$ of $u$, we can find $k\in\mathbb{Z}$ large enough so that $T^k(W) = [0,1]$ because we can accommodate a tent whose base is contained in $W$. \ednote{B: This last phrase doesn't make sense to me. M: The number of tents grow with $k$ and hence this is obvious; I explained in one of our meetings. Does it make sense?} This implies the existence of two points $w_1$ and $w_2$ in $W$ so that $d(T^kw_1,T^kw_2)= 1$, and by the triangle inequality  at least one of the inequalities $d(T^ku,T^kw_1)> \delta$ or $d(T^ku,T^kw_2)> \delta$ hold when $\delta\in (0,1/2)$.  So $T$ has sensitive dependence on initial conditions.  \ednote{M: It is advised not to introduced newlines in a thesis. For the presentation, yes}

@@

  Finally, for every open interval $(a,b)$, the graph of $T^k$ intersects the graph of the identity map on $[0,1]$. This follows from the fact established above that there exists a $k\in\mathbb{N}$ such that $T^k(a,b) = [0,1]$.If the intersection point has coordinates of the form $(p,p)$, the point $p$ is then a fixed point of $T^k$ and therefore also a periodic point of $T$. Since we have found a periodic point in the interval $W$,  the set of periodic points are dense in $T$.
  \ednote{Expand further on reason for density. M: fixed point of $T^k$ is a periodic point, and we have found in an arbitrary interval.}

@@

To show topological similarity or sameness between two metric or topological spaces, one needs to establish a homeomorphism between the two spaces. 
However, in the study of dynamical systems defined on two spaces, establishing a homeomorphism does not indicate that the systems are (physically) related in any way.  For instance, the maps $Tu=u^2$ and $Tu=1-|2u-1|$ defined on $[0,1]$ have totally different behaviour. \ednote{"related" - in what sense am I saying it here? Physically? M: Dynamically}  To find systems that are dynamically similar, one rather needs to establish a kind of dynamical equivalence which we illustrate in a simple commutativity diagram below.
% Giving errors
\begin{equation}  \label{eqn_conjugacy}
%    \[ 
    \psset{arrows=->, arrowinset=0.25, linewidth=0.6pt, nodesep=3pt, labelsep=2pt, rowsep=0.7cm, colsep = 1.1cm, shortput =tablr}
 \everypsbox{\scriptstyle}
 \begin{psmatrix}
U & U\\%
V & V.
 %%%
%  \ncline{1,1}{1,2}^{T} 
%  \ncline{1,1}{2,1} <{\phi}
%   \ncline{2,1}{2,2}^{S}
%  \ncline{1,2}{2,2} > {\phi}
 \end{psmatrix}
% \]
\end{equation}

@@

\begin{Definition} \rm 
  Consider the dynamical systems $T:U\to{U}$, $S:V\to{V}$ and suppose the relationship $\phi \circ T=S\circ \phi$ holds. If $\phi$ is a homeomorphism, then S is said to be conjugate to T and $\phi$ is called a conugacy. If we relax the criterion on $\phi$ and merely require $\phi$ to be continuous and surjective, $\phi$ is then a semi-conjugacy between $T$ and $S$; S is said to be semi-conjugate to T. 
\end{Definition}

\ednote{M: Saying $\phi$ is then a semi-conjugacy between $T$ and $S$ could be ambiguous as it does not specify which system is on the top in the commutativity. So please edit accordingly}

When $S$ is conjugate to $T$, the dynamics of the two systems are in some way ''dynamically equivalent'' and specifically they are in one-to-one correspondence with one another. However when $S$ is semi-conjugate to $T$ with $\phi$ a many-to-one mapping, the dynamics on $V$ provides a coarse-grained description of the dynamics on $U$ \cite{de2013elements}. When $S$ is semi-conjugate to $T$, it is also common to call $S$ a factor of $T$, or conversely that $T$ is an extension of $S$. In essence, an extension (e.g., \cite{de2013elements}) is a larger system capturing all of the important dynamics of its factor.

It is a very hard or nearly an impossible task to establish the existence of a conjugacy or a semi-conjugacy $\phi$ between two systems \cite{devaney2018introduction}. However, one can verify that a function $\phi$ satisfies the commutativity diagram \ref{eqn_conjugacy}.  For example $\phi(x)=\sin(\pi x/2)^2$ is a conjugacy between two systems $Tu=1-|2u-1|$ and $Sv=4v(1-v)$  defined on $U=[0,1]$.

In establishing that two systems are conjugate (semi-conjugate) to one another, we have also shown that one may choose to work with one systems as opposed to the second and still be guaranteed to obtain information on the latter. This is incredibly useful while forecasting the future evolution of dynamical systems.

\chapter{The Learning Problem}

\emph{Introduction...}

Consider a relatively simple learning problem:

Given $(u_0, u_1, \ldots, u_m)$ for $m\in\mathbb{Z}$, a finite segment of an orbit of the map $T$ where $T$ is defined by the update equation $u_{n+1} = Tu_n$, forecast the values $u_{m+1}, u_{m+1}$ where the map $T$ is unknown, given that $u_m$.

A practical example of this would be - given the sequential coordinates of an object moving in space, predict the future positions of that object. However we are very rarely, if at all, presented with a problem where the entire state information is available to us in the form $u_n$ at some time-step $n\in\mathbb{N}$. Consider then a more involved learning problem:

>>>>

\chapter{Introduction}\label{ch1}

@@

The report is organised as follows: 
\newline In Chapter~\ref{ch2}, we recall the definition of a discrete-time dynamical system, how a discrete-system arises from a flow of an ODE and then proceed to define the inverse limit space and topological conjugacy of autonomous systems. 
\newline In Chapter~\ref{ch3}, we introduce the problem of forecasting dynamical systems, state the Takens delay embedding theorem and discuss various issues faced while forecasting. 
\newline In Section 4, a driven dynamical system is defined and discuss the properties of a specific class of driven dynamical systems that we make use of in this thesis/report/paper. 
\newline Finally in Section 5 we show the implementation of these forecasting methods, and conclusions are provided in Section 6.

\chapter{Discrete-time Dynamical Systems} \label{ch2}

@@

Formally, a function $T: U \to U$, where $U$ is some set is  a \emph{discrete-time dynamical system} and its iterates $\{u,Tu,T^2u,\ldots\}$, where $T^n$ denotes the $n$-fold composition of $T$ with itself, describe the evolution of an initial condition $u\in U$ (Note that we frequently drop the brackets and denote $T(u)$ by $Tu$ so as to simplify notations).

Continuous-time dynamical systems modelled using ordinary differential equations can give rise to discrete-time dynamical systems. To see this, consider a differential equation $\dot{x} = f(x)$, $f: \mathbb{R}^n \to \mathbb{R}^n$, $n\in\mathbb{N}$ given to have a unique solution passing through 
each point $x\in\mathbb{R}^{n}$, call it $x(t)$ where $x(0)=x_0$.

\begin{Definition}
  [\bf Flow of an Equation] \label{Dfn_Flow}
  The flow of the equation  $\dot{x} = f(x)$  is defined to be a mapping $\varphi: \mathbb{R}^n \times \mathbb{R} \to \mathbb{R}$ where $\varphi(x_0,t)= x(t)$ for the solution $x(t)$ with $x(0)=x_0$ 
\end{Definition}

By fixing $t=K \in (0,\infty)$, we can define the \emph{time-$K$ map} as  $T(x):= \varphi(x_0,K)$, and it is easily verified that $T\circ T(x_0) = \varphi(x_0,2K)$. In general the $m^{\mbox{th}}$ iterate of $x_0$ under $T$ would be the value of the solution of the ODE evaluated at time $mK$ with the initial condition $x_0$, i.e. $\varphi(x_0, mK)$. 
Thus ordinary differential equations give rise to a discrete-time dynamical system by sampling the value of the solution $x(t)$ at time intervals $K$ units apart.

A numerical discretization of a differential equation can also give rise to a discrete-time dynamical system. For instance, Euler's method approximates $\dot{x}(t)$ by $(x(t+h)-x(t))/h$; if $h$ is fixed throughout, the solution of a differential equation $\dot{x}=f(x)$ 
can be approximated at the time instant $t+(m+1)h$ by iterating the equation 
\begin{equation}
  x(t+(m+1)h) = x(t+mh) + h f(x(t+mh))
\end{equation}

Adopting more succinct notation by replacing $x(t+mh)$ with $u_m$, we rewrite the above equation as
\begin{equation}
u_{m+1} = u_m + hf(u_m)
\end{equation}
or in even simpler terms as the discrete-time dynamical system with map $T(u) = u + hf(u)$, where $T$, $u$ and $f$ are understood to be as above.

Of course, discrete-time dynamical systems need not always arise through a differential equation, and once again we may consider the field of ecology where discrete dynamical systems are often directly derived or assumed. \textbf{Cite.}
There is a school of thought that advocates discrete-time dynamical systems to be more natural for modelling real-world observations than differential equations and we refer the interested reader to \cite{saber2010introduction}.

\section{Invariant Sets}

A core concept in the study of dynamical systems is that of invariance. 
\begin{Definition}
  [\bf Invariant Set] \label{Dfn_Flow}
  Given a discrete-time dynamical system $T: U \to U$, a subset $A \subset U$ is said to be an \emph{invariant set} if $T(A) =A$. 
\end{Definition}

We also define the \emph{orbit of $T$} to be a sequence $\bar{u} = \{u_n\}_{n\in \mathbb{Z}}$ obeying the update equation, $u_{n+1}=Tu_n$, $n \in \mathbb{Z}$.

Two examples of invariant sets include a fixed point where $Tu=u$ for  $u\in U$, and a periodic orbit, i.e., a set of iterates $\{u,Tu, T^2u,\ldots,T^pu\}$, where $T^{p+1}u=u$ for some $p\in\mathbb{Z}$.  The entire space $U$ could be also be invariant.
Consider for example the space $U=[0,1]$, where $Tu=4u(1-u)$ and then $U$ is invariant (as every $u\in{U}$ can be written as $u = 4x(1-x)$ for some $x\in{U}$).

@@

However,  when the space $U$ is compact, all orbits approach an invariant set. This is since the set of limit points of the orbit can be shown to be invariant \cite{de2013elements}. So, when $U$ is compact,  the $\omega$-limit set $\omega(u;T)$ of a point $u$ defined to the collection of limit points of the sequence $\{x,Tu,T^2u,\ldots\}$ is nonempty, and $\omega(u;T)$ is invariant.

\begin{Example}
  For the map $Tu=u^2$ defined on $[0,1]$ all orbits lie in the invariant set $\{1\}$ or else would would approach the invariant set $\{0\}$. 
  % \ednote{B: not relevant other than being an additional example perhaps? M: Have added another simple example}.
\end{Example}

Invariant sets have various properties. Invariant sets can be attracting or repelling depending on how orbits in its vicinity behave. 
Recall the examples $Tu =u/2$ and $Tu=2u$ defined on $\mathbb{R}$ (where $\{0\}$ ``attracts" orbits) or the example, $Tu=u^2$ defined on $[0,1]$ (where $\{1\}$ ``repels" orbits). We are interested in attractive invariant sets since they capture the long-term dynamics as time increases. 
In particular we are interested in those invariant sets named attractors.

\begin{Definition} 
  [\bf Attractor] \label{Dfn_Attractor}
  Let $T: U \to U$, where $U$  is a metric space with metric $d$. A compact subset $A \subset U$ is said to be an attractor if it satisfies the three conditions: 
  \vspace{-8mm}
  \begin{enumerate}
	\item $A$ is invariant. 
	\item $A$ is asymptotically stable, i.e., for every $\epsilon > 0$ and for all $u$ so that $d(u,A) < \epsilon$, we have $d(T^nu,A) \to 0$ as $n\to \infty$. 
	\item $A$ has Lyapunov stability, i.e., for every $\epsilon > 0$  there exists a $\delta(\epsilon) > 0$ so that $d(u,A) < \delta$ implies $d(T^nu,A) < \epsilon$ for all $n\ge 0$.  
\end{enumerate}
\end{Definition}

@@

\section{Chaos}

% \ednote{ should one add paragraphs like these? M: this is a thesis, and not a paper; the instructions sent to the external examiner regarding the thesis evaluation says that the student suggest strongly that it is good to write}
In the 1960s, a number of mathematicians and mathematically interested scientists independently discovered chaos in the mathematical sense. The meteorologist Edward Lorenz may have been the first to explain this phenomenon in his 1963 paper \cite{lorenz1963deterministic}. The notions of invariance, attractivity and chaos may also be described for continuous systems, and Lorenz's system comprised of a system of differential equations. The narrative of Lorenz's discovery of chaos, as well as the history of other forerunners in this subject, are fascinating. For those who are interested, we highly recommend James Gleick's book Chaos: The Making of a New Science \cite{gleick2008chaos}. It clearly illustrates these experiences, explains why chaos was such a startling and crucial mathematical and scientific discovery, and describes the underlying mathematical notions for non-specialists.

There are several intricately different definitions of chaos available in the literature \textbf{(Cite)}, with each of them indicating some aspect of complexity. \ednote%{B:  what are we saying here? M: Devaney's definition is not the only one, there are about half a dozen definitions as well}  
In practice, it is not possible to verify which definition or notion does the underlying dynamical system satisfies when only data is observed from it. We do, however, specifically recall the definition of chaos in the sense of Devaney \cite{devaney2018introduction,de2013elements} so as to understand some nuances behind the complexity. Devaney's definition of chaos has three requirements, with the first being the notion  of sensitive dependence on initial conditions.

\begin{Definition}
  [\bf {Sensitive Dependence on Initial Conditions}] \label{Dfn_SDIC}
A dynamical system $T: U \to U$ is said to have sensitive dependence on initial conditions (SDIC) if there exists a $\delta > 0$ such that for every $u \in U$ and in every neighborhood of $u \in U$ there exists a $v\in{U}$ and an integer $N:=N{(u,v)}\in\mathbb{Z}$ such that $d(T^Nu,T^Nv)>\delta$. 	
\end{Definition}

It is important to acquire a sense of this definition as it can easily be misunderstood. It is common misconception to interpret SDIC as two close points ($u$ and $v$) that eventually become separated by a distance $\delta$ under iteration by $T$. But this is not true. To fully comprehend all the subtleties of the concept, one needs to discuss it more thoroughly by considering each sentence with care and attention, whereafter one may examine how they fit together to convey the concept of sensitive dependence.

@@

To illustrate the concept of SDIC, we present two examples - one from Mathematics and the other from the field of Physics.

\begin{Example}
  The logistic map(LM), a recursion relation of the form $x_{n+1}=rx_n(1-x_n)$ where $x_n\in[0,1]$ and $r<4$ is of particular interest to us as it is classically used to illustrate the influence of chaos on a system \textbf{(Cite)}. Below are plotted the first 75 iterates of the LM for $r=3.7$ and $x_0=0.2$ with an adjustment of 0.001 in $x_0$ and then $r$.

  \begin{figure}[ht]
    \includegraphics[scale=0.45]{log_sdic.eps}
        \centering
        \label{fig:log_sdic}
        \captionof{figure}{Logistic Map for $r=3.7$, $x_0=0.2$}
      \end{figure}

  For an intuitive sense of the fact that the LM exhibits SDIC, we draw two green neighbourhoods of diameters $\delta_1=0.1, \delta_2=0.02$ around the points $u_1=0.7, u_2=0.4$ respectively. It is easy to see that every neighbourhood around these two points will contain some $v_1, v_2\in[0,1]$ and natural number $N_1, N_2 \in\mathbb{N}$ respectively s.t. the distance $d(T^Nu_1,T^Nv)>\delta_1=0.02$ and $d(T^Nu_2,T^Nv)>\delta_2=0.1$.

  In a more rigourous fashion ....
  \ednote{B:Should I be more rigourous here? }

\end{Example}

\begin{Example}
  A second example pertains to a simple physical system called the double pendulum - a pendulum with another pendulum attached to its head given an initial position and angular velociy.

  Below are two figures denoting the trajectories of a double pendulum's second head after some elapsed time. Depicted are three pendulum heads with equal angular velocity, but differing ever so slightly in initial position . As can be seen below, the trajectories diverge very quickly to become completely different. 
  The double pendulum will be discussed in much greater detail in a later chapter (\ref{ch5}), but we mention the example here to provide a numerical example of a practical system exhibiting SDIC.

\begin{figure}[ht]
  \centering
\includegraphics[scale=0.8]{dp_sdic.eps}
    \captionof{figure}{Three double pendulum heads with initial angle differing by 0,025 radians}\label{fig:dp_sdic}
  \end{figure}

\end{Example}

@@

\begin{Definition}
  [\bf {Topologically Transitive}] \label{Dfn_TopTrans}
	A dynamical system $F: U \to U$ is topologically transitive if for any pair of nonempty open sets $E_1$ and $E_2$ there exists a $n\in\mathbb{N}$ such that $T^n(E_1) \cap E_2 \not= \emptyset$. 
\end{Definition}

Topologically transitivity implies that the iterates of an open set of initial conditions gets mixed up with other open sets. On a compact metric space, it can be shown that topological transitivity also implies the existence of a point whose forward iterates are dense \cite{de2013elements}; or in other words, the orbit going through this point will be dense in the compact metric space. 
In fact, it is topologically more likely that the choice of an arbitrary point will be one  whose iterates are almost dense.

\begin{Definition} 
  [\bf {Meager Set}] \label{Dfn_Meager Set}
A subset of a topological space $U$ is said to be a meager set if it can be written as a countable union of sets of with empty interior. The complement of the a meager set is set to be a residual set.
\end{Definition}

@@

We may now define the notion of Chaos as formulated by Devaney\cite{devaney2018introduction}.
\begin{Definition}
  [\bf {Devaney's Chaos}]\label{Dfn_ChaosDec}
	A dynamical system $T: U \to U$ is said to exhibit chaos in the sense of Devaney if it satisfies the three properties:
	\vspace{-5mm}
  \begin{enumerate}
		\item $T$ has SDIC.
		\item $T$ is topologically transitive.
		\item The set of periodic points of $T$ are dense in $U$. 
	\end{enumerate}
\end{Definition}

\begin{Example} 
  The standard tent map $Tu=1-|2u-1|$ defined on $[0,1]$ is a well-known example of a dynamical system satisfying these three properties.

  We now reason as to why this is true. The graph of the map $T$ is piecewise linear with two straight lines, one connecting the points $(0,0)$ and $(\frac{1}{2},1)$ and the other connecting $(\frac{1}{2},1)$ with $(1,0)$. They form a so-called tent with base centered at $u=1/2$. The graph of the map $T^2$ comprises two symmetric tents with their base centered at $1/4$ and $3/4$. See Figure \ref{fig:T2tentmap}

  \begin{figure}[ht]
    \includegraphics[scale=0.6]{T2.png}
        \centering
        \captionof{figure}{Three double pendulum heads with initial angle differing by 0,025 radians}\label{fig:T2tentmap}
  \end{figure}

  In general, the graph of  $T^k$ contains $2^{k-1}$ tents.
  Given any point $u$ and an open neighborhood $W \subset[0,1]$ of $u$, we can find $k\in\mathbb{Z}$ large enough so that $T^k(W) = [0,1]$ because we can accommodate a tent whose base is contained in $W$. % \ednote{B: This last phrase doesn't make sense to me. M: The number of tents grow with $k$ and hence this is obvious; I explained in one of our meetings. Does it make sense?} 
  This implies the existence of two points $w_1$ and $w_2$ in $W$ so that $d(T^kw_1,T^kw_2)= 1$, and by the triangle inequality  at least one of the inequalities $d(T^ku,T^kw_1)> \delta$ or $d(T^ku,T^kw_2)> \delta$ hold when $\delta\in (0,\frac{1}{2})$.  So $T$ has sensitive dependence on initial conditions.

@@

  Finally, for every open interval $(a,b)$, the graph of $T^k$ intersects the graph of the identity map on $[0,1]$. This follows from the fact established above that there exists a $k\in\mathbb{N}$ such that $T^k(a,b) = [0,1]$.If the intersection point has coordinates of the form $(p,p)$, the point $p$ is then a fixed point of $T^k$ and therefore also a periodic point of $T$. Since we have found a periodic point in the interval $W$,  the set of periodic points are dense in $T$.
  This follows because any fixed point of $T^k$ is also a periodic point, and we have found this in an arbitrary interval.% \ednote{Expand further on reason for density. M: fixed point of $T^k$ is a periodic point, and we have found in an arbitrary interval.}

\end{Example}

@@

To show topological similarity or sameness between two metric or topological spaces, one needs to establish a homeomorphism between the two spaces. 
However, in the study of dynamical systems defined on two spaces, establishing a homeomorphism does not indicate that the systems are dynamically related in any way.  For instance, the maps $Tu=u^2$ and $Tu=1-|2u-1|$ defined on $[0,1]$ have totally different behaviour. 
To find systems that are dynamically similar, one rather needs to establish a kind of dynamical equivalence which we illustrate in a simple commutativity diagram below.

% Giving errors
\begin{equation}  \label{eqn_conjugacy}
%    \[ 
    \psset{arrows=->, arrowinset=0.25, linewidth=0.6pt, nodesep=3pt, labelsep=2pt, rowsep=0.7cm, colsep = 1.1cm, shortput =tablr}
 \everypsbox{\scriptstyle}
 \begin{psmatrix}
U & U\\%
V & V.
 %%%
%  \ncline{1,1}{1,2}^{T} 
%  \ncline{1,1}{2,1} <{\phi}
%   \ncline{2,1}{2,2}^{S}
%  \ncline{1,2}{2,2} > {\phi}
 \end{psmatrix}
% \]
\end{equation}

@@

\begin{Definition}  
  [\bf {Conjugacy \& Semi-Conjugacy}]\label{Dfn_Conjugate}
  Consider the dynamical systems $T:U\to{U}$, $S:V\to{V}$ and suppose the relationship $\phi \circ T=S\circ \phi$ holds. If $\phi:U\to{V}$ is a homeomorphism, then $S$ is said to be conjugate to $T$ and $\phi$ is called a conjugacy. If we relax the criterion on $\phi$ and merely require $\phi$ to be continuous and surjective where $\phi:U\to{V}$, then $\phi$ is a semi-conjugacy between $T$ and $S$ where T has domain $U$ and S has domain $V$; S is said to be semi-conjugate to T. 
\end{Definition}

\ednote{M: Saying $\phi$ is then a semi-conjugacy between $T$ and $S$ could be ambiguous as it does not specify which system is on the top in the commutativity. So please edit accordingly. B: Does defining explicitly the domain and codomain of $phi$ make it clear enough? }
When $S$ is conjugate to $T$, the dynamics of the two systems are in some way ''dynamically equivalent''. Specifically, they are in one-to-one correspondence with one another. However when $S$ is semi-conjugate to $T$ with $\phi:U\to{V}$ a many-to-one mapping, the dynamics on $V$ provide merely a coarse-grained description of the dynamics on $U$ \cite{de2013elements}. When $S$ is semi-conjugate to $T$, it is also common to call $S$ a \emph{factor of $T$}, or conversely that $T$ is an \emph{extension of $S$}. In essence, an extension (e.g., \cite{de2013elements}) is a larger system capturing all of the important dynamics of its factor.

It is a very hard or nearly an impossible task to establish the existence of a conjugacy or a semi-conjugacy $\phi$ between two systems \cite{devaney2018introduction}. However, one can verify that a function $\phi$ satisfies the commutativity diagram \ref{eqn_conjugacy}.

\textbf{NB NB FINISH}
\begin{Example}
  For example $\phi(x)=\sin(\frac{\pi}{2}x)^2$ is a conjugacy between two systems $Tu=1-|2u-1|$ and $Sv=4v(1-v)$  defined on $U=[0,1]$.

  To see this, we consider $\phi\circ{T} = \sin(\frac{\pi}{2}(1-|2u-1|))^2$

  and $S\circ\phi = 4\sin(\frac{\pi}{2}x)^{2}(1-\sin(\frac{\pi}{2}x)^2)$
\end{Example}

In establishing that two systems are conjugate (semi-conjugate) to one another, we have also shown that one may choose to work with one system as opposed to the second and still be guaranteed to obtain information on the latter. This is especially useful while forecasting the future evolution of dynamical systems.
